Provably Correct, Low-Cost Optimizations for Abstract Abstract Machines

### 2013-03-18 paper/paper.tex
Finished draft of imperative section.

To describe this process, we'll refactor the transition relation so that it operates on
a pair of a set of contexts ($C$) and a store ($\sigma$).
%
A context includes all non-store components, \emph{e.g.}, the expression, the environment and the stack.
%
The refactored relation, $\widehat{\machstep}$, becomes:
%
\begin{align*}
(C, \msto) &\mathrel{\widehat{\machstep}} (C', \msto') \\
\mbox{where } C' &= \{c' \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto^c), c \in C\} \\
              \msto' &= \bigsqcup\; \{\msto^c \mid \wn(c,\msto)\mathrel{\machstep} \wn(c', \msto^c), c \in C\} \\
\wn(\ev{\mexp, \menv, \mkont}, \msto) &= \ev{\mexp, \menv, \msto, \mkont} \\
\wn(\co{\mval, \mkont}, \msto) &= \co{\mval, \msto, \mkont} \\
\wn(\ap{\mvalx{u}, \mval, \mkont}, \msto) &= \ap{\mvalx{u}, \mval, \msto, \mkont} \\
\wn(\ans{\mval}, \msto) &= \ans{\msto, \mval}
\end{align*}
%
In effect, the new store is computed as the least upper bound of all
stores after stepping. This definition is wasteful, however. It causes
all states found so far to step each iteration, even if they are not
revisited. This has negative performance and precision
consequences. We instead use a frontier-based semantics that
corresponds to the classic ``worklist'' algorithms for analysis.

@@

\paragraph{Correctness}
The correctness of abstract compilation seems obvious, but it has
never before been rigorously proved. What constitutes correctness in
the case of dropped states, anyway? Applying an abstract bytecode's
function does many ``steps'' in one go, at the end of which, the two
semantics line up again (modulo representation of expressions). This
constitutes the use of a notion of stuttering. We therefore prove a
stuttering bisimulation~\cite{clarke-stuttering} between this
semantics and the previous section's to show precision preservation.

The number of transitions that can occur in succession from an
abstract bytecode is roughly bounded by the amount of expression
nesting in the program. The partial order that defines subexpressions
smaller than the expression in which they occur is well-founded, so we
use this observation to drive our use of a well-founded equivalence
bisimulation (WEB). WEBs are equivalent to the notion of a stuttering
bisimulation, but are more amenable to mechanisation (and thus
rigorous proof). They also only require reasoning over one step of the
reduction relation, which makes the proof concise.

@@

Before we prove this setup is indeed a WEB, we need one lemma that
applying an abstract bytecode's function is equal to refining the
corresponding $\ev{}$ state:
%
\begin{lemma}[Compile/commit]
For all $\mcntr,\mexp,\menv,\msto,\mkont$,
$\compile{\mexp}^\mcntr(\menv,r(\msto),r(\mkont)) = r(\ev[^\mcntr]{\mexp,\menv,\msto,\mkont})$.
\end{lemma}
The proof is by induction on $\mexp$.

\begin{theorem}
$B$ is a WEB on $\lmachstep \cup \cmachstep$
\end{theorem}

@@

Correctness here requires a lemma that relates the differences in the compilation functions.
\begin{lemma}
For all $\mexp, \mcntr,\menv,\msto,\msdiff,\mkont,\msto^*,\mkont^*$,
if $\msto \equiv \msto^*$ and $\mkont \equiv \mkont^*$, then
$\dcompile{\mexp}^\mcntr(\menv,\msto,\msdiff,\mkont) = \ttuple{\mastate}{\msdiff'}$ iff
$\compile{\mexp}^\mcntr(\menv,\msto^*,\mkont^*) = \mstate$ such that
$\mastate,\replay(\msdiff',\msto) \equiv \nw(\mstate)$
\end{lemma}
Proof is by induction on $\mexp$.

\begin{theorem}
If $S \equiv S^*, F \equiv F^*, \msto \equiv \msto^*$, then
$(S,F,\msto) \machstep (S',F',\msto')$ iff
$\exists S^*_1, F^*_1, \msto^*_1.
  S' \equiv S^*_1 \wedge F' \equiv F^*_1 \wedge \msto' \equiv \msto^*_1 \wedge
  (S^*,F^*,\msto^*) \dmachstep (S^*_1,F^*_1,\msto^*_1)$
\end{theorem}
Here $\equiv$ means structurally the same, only compiled expressions
are compiled with the updated compilation function that builds store
deltas. The proof follows by equational reasoning and the previous lemma.

@@

There is a subtle loss of precision in Shivers' original time-stamp
technique that we can fix. For a given abstract state, all writes to
the global store need to be delayed until the analysis considers all
branches from that abstract state. This avoids cross-branch pollution
which would not otherwise happen, e.g., when one branch writes to
address $\maddr$ and another branch reads from address
$\maddr$. Fortunately, given our conversion to log-based stores this
change is straightforward.

At this point, we can begin to think of the analysis as an imperative
algorithm with six major components:
%
\begin{itemize}
 \item{$\msto$, the store;}
 \item{{\tt todo}, the workset;}
 \item{{\tt seen}, a map from states to the timestamps at which they were last seen;}
 \item{$\msdiff$, the store changes for the current step;}
 \item{$\changep$, a boolean tracking whether the stepped state contributed a store change; and}
 \item{{\tt T}, the timestamp of the store.}
\end{itemize}

The new $\eval$ calculation is defined in
figure~\ref{fig:imperative}.  To ensure termination, we guard against adding
$c$ to {\tt todo} by the following check:
\begin{align*}\changep \vee {\tt seen}(c) \neq {\tt T}\end{align*}
If it succeeds, {\tt todo} gets $c$ and we update {\tt seen} to map $c$ to {\tt T}.

After all steps complete, we apply $\msdiff$ to $\msto$ imperatively
(with {\tt replay!}) and increase {\tt T} as long as there was at
least one change in $\msdiff$. This logic leads to termination if we
know that each $\ttuple{\maddr}{\mval{s}}$ in $\msdiff$ would change
the value of $\maddr$ in the current store. Thus, we also guard
additions to $\msdiff$ so that only updates that would change the
store are permitted. Each time $\msdiff$ is successfully extended, we
set $\changep$ to true. Before each individual step, we set it
to false.

\begin{figure}
\begin{alltt}
eval(\(e\)) :=
  \(\msto\), todo, seen, T := \(\varnothing\), \(\varnothing\), [], 0
  \(\compile{e}\)(\(\mtcntr, \varnothing, \varnothing, \mtlst, \kmt\))
  while(true):
    if todo = \(\varnothing\): return (keys(seen), \(\msto\))
    else:
      let old := todo
      todo, \(\msdiff\) := \(\varnothing\), \(\mtlst\)
      foreach c \(\in\) old: \(\changep\) := false; c()
      unless \(\msdiff\) = \(\mtlst\): T += 1; replay!(\(\msdiff, \msto\))
\end{alltt}
\caption{Imperative algorithm}
\label{fig:imperative}
\end{figure}

@@

\balance
\bibliographystyle{plain}
\input{paper.bbl}
%\bibliography{local,bibliography}

>>>>

To describe this process, we'll refactor the transition relation so that it operates on
a pair of a set of contexts ($C$) and a store ($\sigma$).
%
A context includes all non-store components, \emph{e.g.}, the expression, the environment and the stack.
%
The refactored relation, $\widehat{\machstep}$, becomes:
%
\begin{align*}
(C, \msto) &\mathrel{\widehat{\machstep}} (C', \msto') \\
\mbox{where } C' &= \{c' \mid \wn(c, \msto) \mathrel{\machstep} \wn(c', \msto^c), c \in C\} \\
              \msto' &= \bigsqcup\; \{\msto^c \mid \wn(c,\msto)\mathrel{\machstep} \wn(c', \msto^c), c \in C\} \\
\wn(\ev{\mexp, \menv, \mkont}, \msto) &= \ev{\mexp, \menv, \msto, \mkont} \\
\wn(\co{\mval, \mkont}, \msto) &= \co{\mval, \msto, \mkont} \\
\wn(\ap{\mvalx{u}, \mval, \mkont}, \msto) &= \ap{\mvalx{u}, \mval, \msto, \mkont} \\
\wn(\ans{\mval}, \msto) &= \ans{\msto, \mval}
\end{align*}
%
In effect, the new store is computed as the least upper bound of all
stores after stepping. This definition is wasteful, however. It causes
all states found so far to step each iteration, even if they are not
revisited. This has negative performance \emph{and} precision
consequences. We instead use a frontier-based semantics that
corresponds to the classic ``worklist'' algorithms for analysis. The
difference is that the store is not modified in-place, but updated
after all frontier states have been processed. This has implications
for the analysis' precision and determinism. Specifically, higher
precision, and is deterministic even if set iteration is not.

@@

\paragraph{Correctness}
The correctness of abstract compilation seems obvious, but it has
never before been rigorously proved. What constitutes correctness in
the case of dropped states, anyway? Applying an abstract bytecode's
function does many ``steps'' in one go, at the end of which, the two
semantics line up again (modulo representation of expressions). This
constitutes the use of a notion of stuttering. We therefore prove a
stuttering bisimulation~\cite{ianjohnson:BCG88} between this
semantics and the previous section's to show precision preservation.

The number of transitions that can occur in succession from an
abstract bytecode is roughly bounded by the amount of expression
nesting in the program. The partial order that defines subexpressions
smaller than the expression in which they occur is well-founded, so we
use this observation to drive our use of a well-founded equivalence
bisimulation (WEB)~\cite{ianjohnson:manolios-diss}. WEBs are equivalent to the notion of a stuttering
bisimulation, but are more amenable to mechanisation (and thus
rigorous proof). They also only require reasoning over one step of the
reduction relation, which makes the proof concise.

@@

Before we prove this setup is indeed a WEB, we need one lemma that
applying an abstract bytecode's function is equal to refining the
corresponding $\ev{}$ state:
%
\begin{lemma}[Compile/commit]\hfill\\
$\compile{\mexp}^\mcntr(\menv,r(\msto),r(\mkont)) = r(\ev[^\mcntr]{\mexp,\menv,\msto,\mkont})$.
\end{lemma}
The proof is by induction on $\mexp$.

\begin{theorem}[Precision preservation]
$B$ is a WEB on $\lmachstep \cup \cmachstep$
\end{theorem}

@@

Correctness here requires a lemma that relates the differences in the compilation functions.
\begin{lemma}[Compile coherence]
For all $\mexp, \mcntr,\menv,\msto,\msdiff,\mkont,\msto^*,\mkont^*$,
if $\msto \equiv \msto^*$ and $\mkont \equiv \mkont^*$, then
$\dcompile{\mexp}^\mcntr(\menv,\msto,\msdiff,\mkont) = \ttuple{\mastate}{\msdiff'}$ iff
$\compile{\mexp}^\mcntr(\menv,\msto^*,\mkont^*) = \mstate$ such that
$\mastate,\replay(\msdiff',\msto) \equiv \nw(\mstate)$
\end{lemma}
Proof is by induction on $\mexp$.

\begin{theorem}[Equivalence]
If $S \equiv S^*, F \equiv F^*, \msto \equiv \msto^*$, then
$(S,F,\msto) \machstep (S',F',\msto')$ iff
$\exists S^*_1, F^*_1, \msto^*_1.
  S' \equiv S^*_1 \wedge F' \equiv F^*_1 \wedge \msto' \equiv \msto^*_1 \wedge
  (S^*,F^*,\msto^*) \dmachstep (S^*_1,F^*_1,\msto^*_1)$
\end{theorem}
Here $\equiv$ means structurally the same, only compiled expressions
are compiled with the updated compilation function that builds store
deltas. The proof follows by equational reasoning and the previous lemma.

@@

There is two subtle losses of precision in Shivers' original
time-stamp technique that we can fix. Firstly, for a given abstract
state, all writes to the global store need to be delayed until the
analysis considers all branches from that abstract state. This avoids
cross-branch pollution which would not otherwise happen, e.g., when
one branch writes to address $\maddr$ and another branch reads from
address $\maddr$. Fortunately, given our conversion to log-based
stores this change is straightforward. Secondly, the common
implementation strategy for timestamps destructively updates each
state's timestamp. This loses \emph{temporal} information about the
contexts a state is visited in, and in what order. Thus that semantics
is not equivalent to the previous section's semantics.

A fix is to keep a stack of the timestamps seen at each state, and
keep a list of stores corresponding with each timestamp. A more
space-efficient representation that also allows an imperative store
would be timestamping each update to the store in the same style as
states. This changes the notion of store equality, but the new
definition is straightforward. For analysis consumers that disregard
temporal information, the space cost in unwarranted, but the
realization that there is precision to be gained opens possibilities.

The following transfer function has several components that can be
destructively updated, and intermediate sets can be elided by adding
to global sets. In fact, the log of store deltas can be removed as
well, by updating the store in-place, and on lookup, using the first
value timestamped $\le$ the current timestamp. We start with the
purely functional view.

\subsubsection{Pure setup for imperative implementation}

The store maps to a stack of timestamped sets of abstract values.

\begin{align*}
\msto \in \Store &= \Addr \to \Valstack \\
\mvalstack \in \Valstack &= (\Timestamp \times \wp(\Storeable))^*
\end{align*}

Here we formally define what lookup and update mean at a given
timestamp.

\begin{align*}
\lookup(\mvalstack,t) &=
  \left\{
    \begin{array}{ll}
      \mval{s} & \text{ if } \mvalstack = \cons{\ttuple{t'}{\mval{s}}}{\mvalstack'}, t' \le t \\
      \mval{s'} & \text{ if } \mvalstack = \cons{\ttuple{t'}{\mval{s}}}{\cons{\ttuple{t''}{\mval{s'}}}{\mvalstack'}}, t' > t
    \end{array}\right. \\
\msto \sqcup_t [\maddr \mapsto \mval{s}] &= \msto[\maddr \mapsto \mvalstack],\changep \\
 (\mvalstack,\changep) &= \msto(\maddr)\sqcup_t \mval{s} \\
\epsilon \sqcup_t \mval{s} &= \ttuple{t}{\mval{s}} \\
\cons{\ttuple{t'}{\mval{s}}}{\mvalstack} \sqcup_t \mval{s'} &= \cons{\ttuple{t'}{\mval{s}\sqcup\mval{s'}}}{\mvalstack},\strue \text{ if } t' > t \\
\mvalstack \sqcup_t \mval{s} &= \cons{\ttuple{t+1}{\mval{s^*}}}{\mvalstack},\strue
           \text{ if } \mval{s'} \neq \mval{s^*} \\
 \mval{s'} &= \lookup(\msto(\maddr),t) \\
 \mval{s^*} &= \mval{s} \sqcup \mval{s'} \\
\mvalstack \sqcup_t \mval{s} &= \mvalstack,\sfalse \text{ otherwise}
\end{align*}

For the purposes of space, we reuse the $\dmachstep$ semantics and change the definition $\replay$ to track if the store has changed, in order to increment the global timestamp.
\begin{align*}
\replay_t(\epsilon,\msto,\changep) &= \msto,\changep \\
\replay_t(\cons{\ttuple{\maddr}{\mval{s}}}{\msdiff},\msto,\changep) &= \replay_t(\msdiff,\msto',\changep \vee \changep') \\
\text{ where } (\msto',\changep') &= \msto\sqcup_t[\maddr \mapsto, \mval{s}] \\
\replayall_t(\varnothing,\msto,\changep) &= \msto,\changep \\
\replayall_t(\setof{\msdiff}\cup\Xi,\msto,\changep) &= \replayall_t(\Xi, \msto', \changep') \\
\text{ where } (\msto',\changep') &= \replay_t(\msdiff,\msto,\changep)
\end{align*}

Because the store has all the temporal information baked into it, we
rephrase the core semantics in terms of a transfer function. The least
fixed-point of this function gives a more compact representation of
the reduction relation of the previous section.

\begin{align*}
\System &= (\State \to {\Timestamp}^*) \times \wp(\State) \times \Store \times \Timestamp \\
{\mathcal F} &: \System \to \System \\
{\mathcal F}(S,F, \msto,t) &= (S',F',\msto', t') \\
\text{ where }
I &= \setof{(\mstate',\msdiff') \mid
       \mstate \in F,
       (\mstate,\msto^*,\epsilon) \dmachstep
       (\mstate',\msdiff')} \\
\msto^* &= \lambda \maddr.\lookup(\msto(\maddr),t) \\
(\msto',\changep) &= \replayall_t(\setof{\msdiff \mid (\_,\msdiff) \in I},\msto,\sfalse) \\
t' &= \left\{\begin{array}{ll} t+1 & \text{ if } \changep \\
              t   & \text{ otherwise}
             \end{array}\right. \\
F' &= \setof{\mstate \mid (\mstate,\_) \in I, \changep \vee S(\mstate) \neq \cons{t}{\_}} \\
S' &= \lambda \mstate. \left\{\begin{array}{ll}
                               \cons{t'}{S(\mstate)} & \text{ if } \mstate \in F' \\
                               S(\mstate) & \text{ otherwise}
                             \end{array}\right.
\end{align*}

\subsubsection{Pure to imperative}

The intermediate data structures of the above transfer function can all be streamlined into globals that are destructively updated. In particular, there are 5 globals:

\begin{enumerate}
\item{$S$: the \emph{seen} set, though made a map for faster membership tests and updates.}
\item{$F$: the \emph{frontier} set, which must be persisent or copied for the iteration through the set to be correct.}
\item{$\msto$: the store, which represents all stores that occur in the reduction semantics.}
\item{$t$: the timestamp, or length of the store chain - all stores that occur in the semantics are totally ordered due to single-threading the store.}
\item{$\changep$: whether there has been a store change stepping all states in $F$.}
\end{enumerate}

The reduction relation would then instead of building store deltas,
update the global store. We would also not view it as a general
relation, but a function that adds all next states to $F$ if they have
not already been seen. At the end of iterating through $F$, $S$ is
updated with the new states at the next timestamp.

% \begin{figure}\label{fig:pure-timestamp}
% \begin{align*}
% t \in \Timestamp &= {\mathbb N} \\
% \msto \in \Store &= \Addr \to (\Timestamp \times \wp(\Storeable))^* \\
% (S,F,\msto,t) &\tmachstep (S \cup S', F', \msto', t') \\
% \text{ where } I &= \{(\mstate,\msdiff) \mid \mstate \in F, \mstate \
% \end{align*}
% \end{figure}

% \begin{figure}
% \begin{alltt}
% eval(\(e\)) :=
%   \(\msto\), todo, seen, T := \(\varnothing\), \(\varnothing\), [], 0
%   \(\compile{e}\)(\(\mtcntr, \varnothing, \varnothing, \mtlst, \kmt\))
%   while(true):
%     if todo = \(\varnothing\): return (keys(seen), \(\msto\))
%     else:
%       let old := todo
%       todo, \(\msdiff\) := \(\varnothing\), \(\mtlst\)
%       foreach c \(\in\) old: \(\changep\) := false; c()
%       unless \(\msdiff\) = \(\mtlst\): T += 1; replay!(\(\msdiff, \msto\))
% \end{alltt}
% \caption{Imperative algorithm}
% \label{fig:imperative}
% \end{figure}

@@

\balance
\bibliographystyle{plain}
%\input{paper.bbl}
\bibliography{local,bibliography}
